\documentclass{beamer}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{float}
\usepackage{tikz}
\usetikzlibrary{trees}
\usepackage{hyperref}
\usepackage{array}

\setbeamertemplate{navigation symbols}{}

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,
    urlcolor=blue
}

\urlstyle{same}

\title{Bash Workshop I: The Basics}
\author{Frederick Yin}
\institute{JITech}
\date{2023}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{wksp}{
    backgroundcolor=\color{backcolour},
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=b,
    keepspaces=true,
    numbers=left,
    numbersep=5pt,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=4
}

\lstset{style=wksp}

\graphicspath{{./img/}}

\AtBeginSection[]{
    \begin{frame}
        \frametitle{Table of Contents}
        \tableofcontents[currentsection]
    \end{frame}
}

\renewcommand{\tt}{\texttt}

\begin{document}
\frame{\titlepage}

\include{00-intro}
\include{01-files}
\include{02-cli}
\include{03-pipes}

\begin{frame}
\frametitle{The End}
\vspace{1cm}
\centering \Huge {Thank You For Coming!}
\end{frame}

\begin{frame}
\frametitle{Credits}
\begin{itemize}
    \item The Free Software Foundation, logo of GNU Bash.
        \url{https://commons.wikimedia.org/wiki/File:Gnu-bash-logo.svg}
    \item Robin Nicholas, Patricia Saunders and The Open Group, logo of UNIX.
        \url{https://commons.wikimedia.org/wiki/File:UNIX_logo.svg}
    \item The Free Software Foundation, logo of GNU.
        \url{https://commons.wikimedia.org/wiki/File:The_GNU_logo.png}
    \item Larry Ewing, logo of Linux.
        \url{https://commons.wikimedia.org/wiki/File:Linux_logo.jpg}
    \item Poul-Henning Kamp, Beastie.
        \url{https://commons.wikimedia.org/wiki/File:Daemon-phk.svg}
    \item ZxxZxxZ. Linux command-line.
        \url{https://commons.wikimedia.org/wiki/File:Linux_command-line._Bash._GNOME_Terminal._screenshot.png}
\end{itemize}
\end{frame}
\end{document}
